---
title: "Imperial College's rodents: a little description"
author: "Irene Espinosa Manjón"
format: html
editor: visual
---

# Context

The data used in this small study were collected by Dr. Malo and his students at Imperial College in July 2022.
By trapping rodents at Silwood Park in tough field campaings, they obtained valuable data of many measurements, but here I will focus on the following: sex, age, anogenital distance, body weight and, the most difficult to achieve, basal metabolic rate.

```{r}
#| warning: false
library(readxl)
data <- read_xlsx("datosirene.xlsx", col_names = T)
names(data)
data <- data[ ,c("pittag", "sp", "sex", "age", "weight", "agd", "bmr")]
str(data)
library(dplyr)
library(tidyverse)
data <- data %>% 
    mutate(pittag = ifelse(pittag == "NA", "UNKNOWN", pittag)) %>% 
    dplyr::filter(sp != "Af") %>% 
    mutate(across(c("pittag", "sp", "sex", "age"), as.factor)) %>% 
    mutate(agd = as.numeric(agd))
str(data)
```

# Analysis and results

## Species, sexes and ages distribution

These data include two different species, with `r nrow(filter(data, sp == "As"))` observations of the wood mouse (*Apodemus sylvaticus*) and `r nrow(filter(data, sp == "Mg"))` of the red vole (*Myodes glareolus*).
As it is shown below in *Figure 1*, on the one hand, we have `r nrow(filter(data, sex == "F"))` females observations, `r nrow(filter(data, sex == "F" & sp == "As"))` of *A.sylvaticus* (As) and `r nrow(filter(data, sex == "F" & sp == "Mg"))` of *M.glareolus* (Mg).
We also have `r nrow(filter(data, sex == "M"))` males observations, `r nrow(filter(data, sex == "M" & sp == "As"))` of *A.s.* and `r nrow(filter(data, sex == "M" & sp == "Mg"))` of *M.g.* On the other hand, there are `r nrow(filter(data, age == "A" & sp == "As"))` adult (A), `r nrow(filter(data, age == "SA" & sp == "As"))` subadult (SA) and `r nrow(filter(data, age == "J" & sp == "As"))` juvenile (J) *A.s.* observations.
*M.g.* has `r nrow(filter(data, age == "A" & sp == "Mg"))` adults, `r nrow(filter(data, age == "SA" & sp == "Mg"))` subadults and `r nrow(filter(data, age == "A" & sp == "Mg"))` juveniles.

```{r}
#| warning: false
library(ggplot2)
theme <- theme(
    axis.line = element_line(linetype = "solid", size = 0.7),
    panel.grid.major = element_line(colour = "gray90",size = 0.4),
    panel.grid.minor = element_line(linetype = "blank"),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 11,colour = "black"),
    panel.background = element_rect(fill = NA)
    )
a <- ggplot(data, aes(x = sex, fill = sp)) +
    geom_bar(position = "dodge") + 
    scale_fill_manual(values = c("As" = "mediumpurple4", "Mg" = "plum3")) + theme +
    theme(legend.position = "none")
b <- ggplot(data, aes(x = age, fill = sp)) +
    geom_bar(position = "dodge") +
    scale_fill_manual(values = c("As" = "mediumpurple4" ,"Mg" = "plum3")) + theme
c <- ggplot(data, aes(x = age, fill = sex)) +
    geom_bar(position = "dodge") +
    scale_fill_manual(values = c("F" = "darkslategray4", "M" = "palegreen3")) + theme
library(patchwork) 
b + a + c + plot_layout(ncol = 2, nrow = 2)
```

**Figure 1.** Distribution of species, sexes and ages.

## Biometric measurements

It seems that adult females of *A.s.* have the highest weight values on average (Table 1; Figure 2) and, in addition, adult males of *A.s.* show the highest anogenital distances (Figure 3).
These results could be related to reproductive status, which I will not discuss here but would be interesting to explore.
On the other side, juvenile males of *A.s* seem to be the most active or stressed and, on the contrary, juvenile males of *M.g.* are the least active, always comparing measurements taken in a non-breeding, post-absorptive and resting, within a thermoneutral zone.
It could be a consequence of one outlier that increases the juvenile *A.s.* average (`r mean(filter(data, sp == "As" & age == "J" & sex == "M")$bmr)`, Table 1; Figure 4) and, also, because of the limited data for some categories, as bmr would be expected to be higher for those smaller individuals (juveniles).

```{r}
#| warning: false
stat <- data %>% group_by(sp, sex, age) %>%  
    summarise(weight.m = mean(weight, na.rm = T), weight.sd = sd(weight, na.rm = T),
              agd.m = mean(agd, na.rm = T), agd.sd = sd(agd, na.rm = T),
              bmr.m = mean(bmr, na.rm = T), bmr.sd = sd(bmr, na.rm = T))
library(kableExtra)
kable(stat) %>% kable_styling(bootstrap_options = c("striped", "hover"))
```

**Table 1.** Means and standard errors of weight, anogenital distance (agd) and basal metabolic rate (bmr) by species (sp), sexes and ages.

```{r}
#| warning: false
d <- ggplot(data, aes(x = sex, y = weight, fill = sp)) +
    geom_boxplot() + 
    scale_fill_manual(values = c("As" = "mediumpurple4", "Mg" = "plum3")) + theme +
    theme(legend.position = "none")
e <- ggplot(data, aes(x = age, y = weight, fill = sp)) +
    geom_boxplot() +
    scale_fill_manual(values = c("As" = "mediumpurple4", "Mg" = "plum3")) + theme +
    theme(legend.position = "right", axis.title.y = element_blank())
d + e + plot_layout(ncol = 2, nrow = 1, widths = c(1.5, 2))
```

**Figure 2.** Distribution of weight by sexes, ages and species (sp).

```{r}
#| warning: false
f <- ggplot(data, aes(x = sex, y = agd, fill = sp)) +
    geom_boxplot() + 
    scale_fill_manual(values = c("As" = "darkslategray4", "Mg" = "palegreen3")) + theme +
    theme(legend.position = "none")
g <- ggplot(data, aes(x = age, y = agd, fill = sp)) +
    geom_boxplot() +
    scale_fill_manual(values = c("As" = "darkslategray4", "Mg" = "palegreen3")) + theme +
    theme(legend.position = "right", axis.title.y = element_blank())
f + g + plot_layout(ncol = 2, nrow = 1, widths = c(1.5, 2))
```

**Figure 3.** Distribution of anogenital distance (agd) by sexes, ages and species (sp).

```{r}
#| warning: false
h <- ggplot(data, aes(x = sex, y = bmr, fill = sp)) +
    geom_boxplot() + 
    scale_fill_manual(values = c("As" = "lightsalmon4", "Mg" = "lightsalmon1")) + theme +
    theme(legend.position = "none")
i <- ggplot(data, aes(x = age, y = bmr, fill = sp)) +
    geom_boxplot() +
    scale_fill_manual(values = c("As" = "lightsalmon4", "Mg" = "lightsalmon1")) + theme +
    theme(legend.position = "right", axis.title.y = element_blank())
h + i + plot_layout(ncol = 2, nrow = 1, widths = c(1.5, 2))
```

**Figure 4.** Distribution of basal metabolic rate (bmr) by sexes, ages and species (sp).

# Big conclussion

This way to manage, transform and graphic data is fantastic!
Thank you so much for sharing it with us.
